---
title: "AR(1) Probit Model for US Recession Prediction"
author: "MS"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

## Setup

Load required packages and source estimation functions:

```{r setup}
library(tidyverse)
library(lubridate)
library(rsample)

# Source MCL estimation functions
source("R/functions.R")

# Set theme for plots
theme_set(theme_minimal(base_size = 12))
```

## Load and Explore Data

```{r load-data}
# Load quarterly recession data
load("data/df_quarterly.Rdata")

# Overview
df_quarterly %>%
  glimpse()

# Recession summary
df_quarterly %>%
  count(recession_flag) %>%
  mutate(
    status = if_else(recession_flag == 1, "Recession", "Expansion"),
    proportion = scales::percent(n / sum(n), accuracy = 0.1)
  )
```

### Time Series Visualization

```{r plot-timeseries, fig.width=10, fig.height=8, echo = F}
# Long format for plotting
df_long <- df_quarterly %>%
  pivot_longer(
    cols = c(yield_spread, NFCI),
    names_to = "variable",
    values_to = "value"
  )

# Plot with recession shading
ggplot() +
  geom_rect(
    data = df_quarterly %>% filter(recession_flag == 1),
    aes(xmin = date, xmax = date + months(3), ymin = -Inf, ymax = Inf),
    fill = "gray80", alpha = 0.5
  ) +
  geom_line(
    data = df_long,
    aes(x = date, y = value, color = variable),
    linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~variable, ncol = 1, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "US Recession Predictors Over Time",
    subtitle = "Gray bars indicate recession quarters",
    x = "Date",
    y = "Value"
  ) +
  theme(legend.position = "none")
```

## Time Series Cross-Validation Setup

Create expanding window resamples for out-of-sample evaluation:

```{r create-folds}
# Create expanding window resamples
# Initial training: 40 quarters (10 years)
# Step forward 1 quarter at a time
# This creates a TRUE rolling forecast: each window fits at time T and forecasts T+1, T+2, etc.

folds <- sliding_period(
  df_quarterly,
  index = date,
  period = "quarter",
  lookback = Inf,         # Expanding window (use all history)
  assess_stop = 4,        # Assess next 4 quarters for h=1,2,3,4 evaluation
  skip = 39,              # Skip first 39 quarters (need 40 for initial training)
  step = 1                # Step forward 1 quarter
)

# Inspect the folds
cat("Number of resamples:", nrow(folds), "\n\n")

cat("First 3 folds:\n")
for (i in 1:min(3, nrow(folds))) {
  split <- folds$splits[[i]]
  train_data <- analysis(split)
  test_data <- assessment(split)

  cat("Fold", i, ":\n")
  cat("  Training (up to T): n =", nrow(train_data),
      " | ", as.character(min(train_data$date)), "to",
      as.character(max(train_data$date)), "\n")
  cat("  Testing (T+1 to T+4): n =", nrow(test_data),
      " | ", as.character(min(test_data$date)), "to",
      as.character(max(test_data$date)), "\n\n")
}
```

### Visualize Resampling Scheme

```{r plot-folds, fig.width=10, fig.height=8, echo = F}
# Extract fold information for visualization
fold_info <- map_dfr(seq_len(min(30, nrow(folds))), function(i) {
  split <- folds$splits[[i]]
  train_data <- analysis(split)
  test_data <- assessment(split)

  tibble(
    fold = i,
    train_start = min(train_data$date),
    train_end = max(train_data$date),
    test_start = min(test_data$date),
    test_end = max(test_data$date)
  )
})

# Plot (show first 30 folds for clarity)
fold_info %>%
  ggplot(aes(y = fold)) +
  geom_segment(
    aes(x = train_start, xend = train_end, yend = fold),
    color = "steelblue", linewidth = 2
  ) +
  geom_segment(
    aes(x = test_start, xend = test_end, yend = fold),
    color = "firebrick", linewidth = 2
  ) +
  scale_y_reverse() +
  labs(
    title = "Expanding Window Resampling Scheme (First 30 Folds)",
    subtitle = "Blue = Training (up to time T) | Red = Testing (T+1 to T+4)",
    x = "Date",
    y = "Fold"
  ) +
  theme(panel.grid.minor = element_blank())
```

## Model Estimation and Forecasting Across Windows

For each window:
1. Fit model on training data (up to time T)
2. Generate forecasts for h=1,2,3,4 using ONLY information at time T
3. Compare forecasts to actual values in test period

```{r rolling-forecasts}
# Function to evaluate one fold
evaluate_fold <- function(split, fold_id) {

  # Extract train and test data
  train_data <- analysis(split)
  test_data <- assessment(split)

  # Fit model on training data (up to time T)
  fit <- estimate_ar_probit_mcl(
    recession_flag ~ yield_spread + NFCI,
    data = train_data,
    lag_m = 4
  )

  # Generate forecasts for h=1,2,3,4 from time T
  # NO test data is used in prediction - only training info
  horizons <- 1:4
  forecasts <- predict(fit, h = horizons, type = "prob")

  # Create results tibble
  # Match forecasts to actual values
  results <- tibble(
    fold = fold_id,
    train_end = max(train_data$date),
    horizon = horizons,
    forecast_date = train_end + months(3 * horizons),
    forecast_prob = forecasts,
    actual = NA_real_
  )

  # Match actual values from test data
  for (j in 1:nrow(results)) {
    matching_row <- which(test_data$date == results$forecast_date[j])
    if (length(matching_row) > 0) {
      results$actual[j] <- test_data$recession_flag[matching_row[1]]
    }
  }

  # Add coefficient estimates
  results$rho <- fit$coefficients["rho"]
  results$alpha <- fit$coefficients["alpha"]
  results$beta_yield <- fit$coefficients["yield_spread"]
  results$beta_nfci <- fit$coefficients["NFCI"]
  results$convergence <- fit$convergence

  return(results)
}

# Evaluate all folds (this may take a few minutes)
cat("Evaluating", nrow(folds), "folds...\n")

all_forecasts <- map_dfr(
  seq_len(nrow(folds)),
  function(i) {
    if (i %% 20 == 0) cat("  Fold", i, "of", nrow(folds), "\n")
    evaluate_fold(folds$splits[[i]], i)
  }
)

# Remove rows with missing actual values (happens at the end)
all_forecasts <- all_forecasts %>%
  filter(!is.na(actual))

cat("Done! Evaluated", length(unique(all_forecasts$fold)), "folds.\n")
cat("Total forecasts:", nrow(all_forecasts), "\n")
```

## Forecast Performance Metrics

Calculate performance metrics by horizon:

```{r metrics}
# Calculate metrics by horizon
metrics_by_horizon <- all_forecasts %>%
  group_by(horizon) %>%
  summarise(
    n_forecasts = n(),
    mean_forecast = mean(forecast_prob),
    mean_actual = mean(actual),
    # Mean Squared Error (Brier Score)
    mse = mean((forecast_prob - actual)^2),
    # Mean Absolute Error
    mae = mean(abs(forecast_prob - actual)),
    # Log Loss
    log_loss = -mean(actual * log(pmax(forecast_prob, 1e-15)) +
                     (1 - actual) * log(pmax(1 - forecast_prob, 1e-15))),
    .groups = "drop"
  )

metrics_by_horizon
```

### Performance by Horizon

```{r plot-metrics-horizon, fig.width=10, fig.height=6}
# Reshape for plotting
metrics_long <- metrics_by_horizon %>%
  select(horizon, mse, mae, log_loss) %>%
  pivot_longer(
    cols = c(mse, mae, log_loss),
    names_to = "metric",
    values_to = "value"
  )

# Plot
metrics_long %>%
  ggplot(aes(x = horizon, y = value, color = metric)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = 1:4) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Forecast Performance by Horizon",
    subtitle = "Performance degrades slightly with longer horizons",
    x = "Forecast Horizon (quarters ahead)",
    y = "Metric Value"
  ) +
  theme(legend.position = "none")
```

### Metrics Over Time (for h=1)

```{r metrics-over-time, fig.width=10, fig.height=8}
# Calculate rolling metrics for h=1 forecasts
h1_forecasts <- all_forecasts %>%
  filter(horizon == 1) %>%
  arrange(forecast_date) %>%
  mutate(
    squared_error = (forecast_prob - actual)^2,
    abs_error = abs(forecast_prob - actual),
    log_loss_contrib = -(actual * log(pmax(forecast_prob, 1e-15)) +
                        (1 - actual) * log(pmax(1 - forecast_prob, 1e-15)))
  )

# Plot metrics over time
h1_forecasts %>%
  select(forecast_date, squared_error, abs_error, log_loss_contrib) %>%
  pivot_longer(
    cols = c(squared_error, abs_error, log_loss_contrib),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = forecast_date, y = value)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "firebrick", linewidth = 0.8) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(
    title = "1-Quarter Ahead Forecast Errors Over Time",
    subtitle = "Blue line = actual errors | Red line = smoothed trend",
    x = "Forecast Date",
    y = "Error"
  )
```

## Forecast Visualization

### Out-of-Sample Forecasts (h=1)

```{r plot-forecasts-h1, fig.width=10, fig.height=6}
# Plot 1-quarter ahead forecasts
h1_forecasts %>%
  ggplot(aes(x = forecast_date)) +
  # Recession shading
  geom_rect(
    data = df_quarterly %>% filter(recession_flag == 1),
    aes(xmin = date, xmax = date + months(3), ymin = -Inf, ymax = Inf),
    fill = "gray80", alpha = 0.5, inherit.aes = FALSE
  ) +
  # Forecast probabilities
  geom_line(aes(y = forecast_prob), color = "steelblue", linewidth = 1) +
  geom_point(
    data = h1_forecasts %>% filter(actual == 1),
    aes(y = forecast_prob), color = "firebrick", size = 2
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title = "Out-of-Sample 1-Quarter Ahead Recession Forecasts",
    subtitle = "Gray bars = actual recessions | Red points = recession quarters | Blue line = forecasted probability",
    x = "Forecast Date",
    y = "P(Recession)"
  )
```

### Comparison Across Horizons

```{r plot-all-horizons, fig.width=10, fig.height=6}
# Plot all horizons for recent period
recent_forecasts <- all_forecasts %>%
  filter(forecast_date >= as.Date("2000-01-01"))

recent_forecasts %>%
  ggplot(aes(x = forecast_date, y = forecast_prob, color = factor(horizon))) +
  # Recession shading
  geom_rect(
    data = df_quarterly %>%
      filter(recession_flag == 1, date >= as.Date("2000-01-01")),
    aes(xmin = date, xmax = date + months(3), ymin = -Inf, ymax = Inf),
    fill = "gray80", alpha = 0.3, inherit.aes = FALSE
  ) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_brewer(palette = "Set1", name = "Horizon") +
  labs(
    title = "Out-of-Sample Forecasts by Horizon (2000-Present)",
    subtitle = "Gray bars indicate actual recessions",
    x = "Forecast Date",
    y = "P(Recession)"
  )
```

### Calibration Plot

```{r calibration, fig.width=7, fig.height=6}
# Calibration plot for h=1 forecasts
h1_forecasts %>%
  mutate(
    prob_bin = cut(
      forecast_prob,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_predicted = mean(forecast_prob),
    mean_actual = mean(actual),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(prob_bin)) %>%
  ggplot(aes(x = mean_predicted, y = mean_actual)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(size = n), color = "steelblue", alpha = 0.7) +
  geom_line(color = "steelblue") +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title = "Calibration Plot (1-Quarter Ahead, Out-of-Sample)",
    subtitle = "Perfect calibration lies on the diagonal",
    x = "Predicted Probability",
    y = "Observed Frequency",
    size = "n obs"
  ) +
  coord_equal()
```

## Parameter Stability

Check if estimated parameters are stable across windows:

```{r parameter-stability, fig.width=10, fig.height=8}
# Parameter evolution over time
param_evolution <- all_forecasts %>%
  filter(horizon == 1) %>%  # One row per fold
  select(fold, train_end, rho, alpha, beta_yield, beta_nfci) %>%
  distinct()

# Plot
param_evolution %>%
  pivot_longer(
    cols = c(rho, alpha, beta_yield, beta_nfci),
    names_to = "parameter",
    values_to = "value"
  ) %>%
  ggplot(aes(x = train_end, y = value)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "firebrick", linewidth = 0.6) +
  facet_wrap(~parameter, scales = "free_y", ncol = 2) +
  labs(
    title = "Parameter Estimates Over Time",
    subtitle = "Blue line = estimates | Red line = smoothed trend",
    x = "Training Period End Date",
    y = "Parameter Value"
  )
```

## Final Model

Fit final model on all available data:

```{r final-model}
# Fit on full dataset
final_fit <- estimate_ar_probit_mcl(
  recession_flag ~ yield_spread + NFCI,
  data = df_quarterly,
  lag_m = 4
)

# Display results
summary(final_fit)
```

### In-Sample Fit

```{r final-fit-plot, fig.width=10, fig.height=5}
# Get fitted probabilities
fitted_probs <- fitted(final_fit, type = "prob")

# Create data frame (accounting for lag)
fitted_df <- tibble(
  date = df_quarterly$date[(final_fit$lag_m + 1):nrow(df_quarterly)],
  actual = final_fit$y,
  fitted_prob = fitted_probs
)

# Plot
fitted_df %>%
  ggplot(aes(x = date)) +
  geom_rect(
    data = fitted_df %>% filter(actual == 1),
    aes(xmin = date, xmax = date + months(3), ymin = -Inf, ymax = Inf),
    fill = "gray80", alpha = 0.5
  ) +
  geom_line(aes(y = fitted_prob), color = "steelblue", linewidth = 1) +
  geom_point(
    data = fitted_df %>% filter(actual == 1),
    aes(y = fitted_prob), color = "firebrick", size = 2
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title = "Final Model: In-Sample Fitted Probabilities",
    subtitle = "Gray bars = actual recessions | Red points = recession quarters",
    x = "Date",
    y = "P(Recession)"
  )
```
